# Programming 101{#CH2}

> The goal of this chapter is not to turn you into an R programmer, but to cover the most important programming concepts so that you feel more comfortable using R and are able to get up and running with your analyses as fast as possible.

## Running R Code {-}
The rest of this booklet will contain code blocks that look like this:

```{r}
# This is a code block, and this is an example of a comment.
# You can comment your script by using the hashtag.
# That way you can make notes to yourself and keep track of what your code 
# is doing at each step.
```

You can copy and paste the code within the blocks to your R session to try it out for yourself; but the code above is just a comment so it won't do anything exciting! The code blocks will often (but not always) have some sort of output printed directly below them so that you can see the results. For example:

```{r}
mean(1:10)
```

In the example above, the code block consisted of code that calculated the mean of the numbers 1 through 10, and the results were printed directly below. Notice how the output is printed in black font, which is how you can distinguish the code blocks from the output blocks.

You can "run" a line of code in R by pressing **Cmd + Enter** on a Mac and **Ctr + Enter** on a Windows. For example, if you wanted to know what 75 times 82 is, you could type `75 * 82` into your script and "run" the code by using **Cmd + Enter**  or **Ctr + Enter** to get the answer. Or, if you're typing code directly into the console, you can run the code by hitting **Enter**.

## R is a calculator {-}
Try out R's calculator functionality by copying and pasting the code below into your R session.

```{r eval=FALSE}
21 + 21   # Addition
43 - 20   # Subtraction
4 * 4     # Multiplication
45 / 3    # Division
2 ^ 3     # Exponentiation
sqrt(16)  # Square root

# R follows PEMDAS rules
(((3 + 2)^2) + (10 * 2)) - 4
(3 + 2)^2 + 10 * 2 - 4
```

## Objects and Assignment {-}
One of the most basic programming concepts is **assignment**. Let's say we want the letter `x` to refer to the value 5. You can do this by *assigning* `x` to the value 5 with the assignment operator, `<-`, or with the equal sign operator, `=`:

```{r}
x <- 5
x = 5
```

In both of these examples the result is the same: `x` is now assigned to the value 5. It's a matter of personal preference which operator you'd like to use (`<-` or `=`), but it's best to be consistent to make your script more readable. 

`x` is now considered an **object**. An object is a name used to reference a value or set of values. The object `x` refers to the value 5. Instead of just assigning the number 5 to `x`, we could perform several calculations and then assign the result to an object, like so:

```{r}
x <- sqrt(34 + ((12 * 2)^3 - 5))
x # typing x by itself will print the value to the console
```

**The object `x` no longer refers to 5**. We overwrote `x` and assigned it to a new value, ~117.7. 

When analyzing data, you'll likely need to transform the data before you can run  statistical tests. You can accomplish this by transforming the data and then saving the data as a new object. Let's look at one more example.

```{r}
head(iris)
# The head() function prints the first few observations from the dataset
```

Here's one of R's built-in datasets, which is called `iris` and consists of 3 flower species (`setosa`, `virginica`, `versicolor`) and their their petal characteristics. Let's say we only wanted to include `virginica` flowers in our analysis.

```{r eval=FALSE}
subset(iris, Species == "virginica")
```

The code above filters the data to include only `virginica` plants, but we didn't assign it to an object so we have no way of referencing our filtered data.

```{r}
iris_virginica <- subset(iris, Species == "virginica")
head(iris_virginica)
```

The code above looks much better. The filtered data has been assigned to the object `iris_virginica`, so we now have a way to reference that data, and we could use this object to analyze the virginica plants exclusively.

## Dollar Signs and Brackets {-}

## Variable Types {-}
R has several different variable types, and to manipulate your data it is good to be aware of them.

```{r, eval=FALSE}
0.5, 2         # Numeric class
TRUE, FALSE    # Logical class
"hat", "dog"   # Character class
5L, 12L        # Integer class ("L" stores the value as an integer)
1+0i, 2+4i     # Complex class
"yes", "no"    # Factor class (categorical variable; each level is a category)
```

These variable types are the basic building blocks, and all values in R belong to one of them. You may have a situation where your data is imported as `character` but you'd like to change it to `factor`, or vice versa. You can easily do this with the `as.factor()` and `as.character()` functions, which we'll talk more about in Chapter 5. 

```{r}
y <- c(1,2,3,4,5)
y
```

When you add multiple values together, it's called concatenation. In the above example, `numeric` values were concatenated together and assigned to the object `y` The object `y` now refers to 5 values: `1, 2, 3, 4, 5`. When we concatenate values together like this, we create a `data type`.

## Data Types {-}
R has many data types, but the only one we'll worry about is the `data frame`. A `data frame` is usually the object that's created when importing a data file into R. A data frame consists of columns and rows, where each column is the same length. Here's an example of one of R's built-in data frames called `mtcars`:

```{r}
head(mtcars)
```

## Functions and Arguments {-}
So far we've seen R as a calculator, created objects, and learned about variable types and data types. But how we we actually *do* things in R? That is, how do we manipulate data and run a statistical analysis? 

To do just about anything in R you use functions. What functions do in R is take a process that would normally be a hassle for the user and make it easier. For example, the `mean()` function. We're all familiar with how the mean is calculated, and we *could* calculate this manually in R. If we wanted the mean of the numbers `c(1, 3, 4, 4)` we could simply sum the numbers and divide by n to get 3. But why do that when we can just use the `mean()` function? And that's the basic idea of a function: to simplify things for us. The `lm()` function creates a linear model from our data, which again we *could* achieve the same result by manually performing *many* calculations, but this would be a time-consuming, arduous task, and using the `lm()` function greatly simplifies the process. 

Functions are great because they can simplify and streamline a process, but how do you make the function do exactly what *you* need to do in your analysis? Let's take a look at the `factor()` function to get an indea of how to take advantage of functions and arguments. 

Factors are categorical variables, where each level is a category. Here is an example of character data that's been converted to a factor with the `factor()` function:

```{r}
y <- c("cat", "dog", "cat", "cat", "dog", "cat", "dog")
y <- factor(y) # y is converted to a factor with 2 levels
y
```

In this example, two levels are created from the data: `cat` and `dog`. Let's slighly modify this example and say that data was entered incorrectly into a column, so the column looks like this:

```{r}
y <- c("cat", "dg", "cat", "cart", "dog", "cat", "dog")
y <- factor(y) 
y
```

Because of the typos, we now have 4 levels--`cat`, `dg`, `cart`, `dog`-- and that's clearly not what we want. To make the most of the `factor()` function, we can make use of its **arguments**, which allows us to have greater control over what the function does and how it manipulates the data. We could ignore the typos by using the `levels` argument, which is a way of specifying how many levels we're supposed to have for the data:

```{r}
y <- c("cat", "dg", "cat", "cart", "dog", "cat", "dog")
y <- factor(y, levels = c("cat", "dog"))
y
```

We're now back to two levels, and the typos were treated as `NA`. Now let's say we wanted to recode `cat` as `0` and `dog` as `1`. We could use the `labels` argument, which creates a new label for the corresponding level (in order): 

```{r}
y <- c("cat", "dg", "cat", "cart", "dog", "cat", "dog")
y <- factor(y, levels = c("cat", "dog"), labels = c(0, 1))
y
```

By using `factor()`'s arguments we were able to get the function to do what *we* wanted it to do, not just what it does by default. This is what makes R so powerful!

### Example {-}
Now let's look at a more complete example of functions and arguments with the `lm()` function.

```{r, eval=FALSE}
lm(formula, data, subset, weights, na.action,
   method = "qr", model = TRUE, x = FALSE, y = FALSE, qr = TRUE,
   singular.ok = TRUE, contrasts = NULL, offset, ...)
```

Above is the `lm()` function used for fitting linear models. The function has many arguments, but most of the arguments have default values; what this means is that you do not need to assign a value to that argument in order for the function to work. For example, in this function the default `method` argument is set to `qr`: `method = "qr"`. This means that when you use the `lm()` function, the `method` will be set to `qr` even if you don't explicitly set it to `qr`; hence, `qr` is the default (which is referring to [QR Decomposition](https://en.wikipedia.org/wiki/QR_decomposition), in case you were curious).

In addition to arguments that have default values, there are also arguments which are optional. In the `lm()` function above, `subset`, `weights`, `na.action`, and `offset` are all optional arguments. Unsurpringly, this means that these arguments do not *need* to be used, but they can be. 

We can now see why the code below is valid, even though there are many arguments that we didn't specify.

```{r eval=FALSE}
lm(formula = mpg ~ hp, data = mtcars)
```

We can also see why this code below is valid. The `subset` argument is optional, and here we used it to specify that we only want to include the first 10 observations in our model.

```{r eval=FALSE}
lm(formula = mpg ~ hp, data = mtcars, subset = c(1:10))
```

> *But how do I know what the arguments are for a given function, what they do, which arguments have default values, which ones are optional, and which ones I need to specify?*

This is where the documentation comes in handy. You can type a question mark in front of the function you'd like to learn more about (`?factor`). If you're using RStudio, this will bring up the documentation page for that function under the help tab in the lower right pane, which will usually include a description, the arguments (and their descriptions), and example code showing how to use the function. You can also type the function into the search bar under the help tab on the lower right pane.

Hopefully, though, we'll provide enough example code throughout this book that you won't need to concern yourself with all of these details. But, we can't provide examples of *everything* for *every* situation, and you'll almost certainly use functions that aren't covered in this book. This is where having a basic understanding of these topics will come in handy.

## Code Formatting {-}
As a final note, just know that the spacing in your code doesn't matter. You can type `5 * 5` or `5*5` and get the same result. You can type `mean(1:10)` or `mean(1 : 10)` and you'll get the same result. You can even type `mean (1:10)` or `mean (1 : 10)`, but you probably shouldn't because this is very unconventional. 

<!--
## Test Your Knowledge {-}
-->
