# Packages and Libraries

## What are Packages? {-}
R has hundreds of useful functions, but sometimes you'll want or need to use a function that's not built into R. For example, R has several built-in functions for plotting data, such as the `hist()` function:

```{r include=FALSE}
set.seed(1)
```

```{r out.height= '70%', out.width= '70%'}
hist(rnorm(100))
```

And that works fine if you want to quickly analyze your data, but that graph is certainly not publication quality. You could spruce up the plot by installing a **package** that contains functions for creating much more aesthetically pleasing plots, such as the `ggplot2` package.

```{r out.height= '70%', out.width= '70%'}
library(ggplot2)
set.seed(1)
ggplot(tibble::tibble(rnorm = rnorm(100, mean = -1)), aes(x = rnorm)) +
    geom_density(tibble::tibble(rnorm = rnorm(100, mean = 1)), 
                 mapping = aes(x = rnorm), 
                 fill = "blue", 
                 alpha = .7) +
    geom_density(fill = "red", alpha = .7) + 
    theme_bw() +
    labs(title = "A Colorful Histogram Example") + 
    xlab("Random Normal Distribution") +
    ylab("Density") +
    theme(panel.grid.minor.y = element_blank(),
          panel.grid.minor.x = element_blank(),
          panel.grid.major.y = element_blank(),
          panel.grid.major.x = element_blank(),
          text = element_text(size = 16, family = "serif"))
```

The `ggplot2` package is simply a collection of R code and functions, created by an R user, that allows you to make highly customizable plots like the one above without too much difficulty. 

## Installing Packages {-}
There are many packages that you can install to make data analysis easier. One package that we'll use frequently throughout this book is the `tidyverse` package. The `tidyverse` package is actually a collection of packages bundled together-- a meta-package. You can install the `tidyverse` with the following line of code:

```{r eval=FALSE}
install.packages("tidyverse")
# You only need to install a package once
```

## What's a Library? {-}
Once you install a package it becomes part of your "library". Your library is a collection of all the packages that are installed on your computer. If you click on the "packages" tab in the lower right pane, you can see a list of all the packages that are currently installed. This is your library. Notice how the header says "System Library". 

## Loading Packages {-}
Packages only need to be *installed* once, but they need to be *loaded* every time you start an R session. Loading a package puts it into the working memory; that way R knows you want to use those functions from those packages. You can load a package with the `library()` function:

```{r eval=FALSE}
library(tidyverse)
```

Running the code `library(tidyverse)` now allows you to use all of the functions that come bundled with the `tidyverse` package (which is a lot!). 

* Pro tip: It's best practice to load your desired package with the `library()` function at the very beginning of your R script; that way it's loaded *before* the rest of your code is ran.

You can check to see if you installed the `tidyverse` package and loaded it into your R session correctly by running the `ggplot` histogram code above. If you see the same plot in your R session, everything is working correctly!

You might be wondering why the ggplot code worked when we loaded the `tidyverse` package and not the `ggplot2` package. This is because the `tidyverse` is a *meta-package*, which *includes* the `ggplot2` package within it (along with 7 other packages: `dplyr`, `tidyr`, `readr`, `purr`, `tibble`, `stringr`, `forcats`). So, if you get into the habit of loading the tidyverse meta-package at the beginning of your script, that will generally take care of everything for you, and you won't need to worry about loading many (if any) other packages into your R session.
