```{r include=FALSE}
library(tidyr)
library(magrittr)
library(dplyr)
```

# Formatting Data{#CH6}

> This chapter will cover: missing data, merging data files, changing variable types, recoding varibles, data transformation and tidy data principles. 

> The code in this chapter only works if you're following along with the Github folder for this book (which you can download [here](https://github.com/JasonDude16/Statistical-Analysis-Folder)), correctly set your working directory to the data folder (which you can learn how to do in [Chapter 4](#CH4)), and run the code in the order it appears in this chapter. 

## Merging Files {-}
Sometimes you might want to import multiple files and combine them into one dataset. This can be accomplished with the `cbind()` and `rbind()` functions, which are short for column bind and row bind. Let's say you have three files that have the same column names and you want to combine them. 

```{r}
data1 <- read.csv("BMI_1.csv")
dim(data1)

data2 <- read.csv("BMI_2.csv")
dim(data2)

data3 <- read.csv("BMI_3.csv")
dim(data3)

newData <- rbind(data1, data2, data3)
dim(newData)
```

As you can see the last printout shows the dimensions are 60 x 2, which means the three datasets were successfully combined.

Let's say you have two files with *different* column names and want to combine them side-by-side. For example, you might have a data file that contains `subID`, `age` and `sex`, and another file contains `height` and `weight` data. These columns can be added into one dataset with the `cbind()` function. 

`subID`, `age`, and `sex` dataset:

```{r}
data1 <- read.csv("SubID_Age_Sex.csv")
head(data1)
```

`height` and `weight` dataset:

```{r}
data2 <- read.csv("BMI_1.csv")
head(data2)
```

Combind with the `cbind()` function:

```{r}
newData <- cbind(data1, data2)
head(newData)
```

## Long Format {-}
Putting data in long format increases the number of rows and decreases the number of columns in a data frame. In the health sciences, this often means that one row corresponds to one observation (at one point in time). If your data is not in long format, you can change it to long format with the `pivot_longer()` function, which comes from the tidyverse package.

Here's an example of a research study where subjects had their VO2 max measured at four points in time, and it's currently *not* in long format:

```{r}
data_VO2 <- read.csv("data_VO2.csv")
head(data_VO2)
```

We can convert the **data_VO2.csv** into long format with the `pivot_longer()` function:

```{r}
# Make sure the tidyverse package is loaded 
data_VO2_long <- pivot_longer(data = data_VO2, 
                              cols = c("Time1", "Time2", "Time3", "Time4"),
                              names_to = "Time", 
                              values_to = "VO2_max")
head(data_VO2_long)
```

## Wide Format {-}
Putting data in wide format increases the number of columns in a data frame. In the health sciences, this often means that one row will contain multiple points in time; this is how the **data_VO2.csv** was formatted before we converted it to long format. We can now change it back into wide format with the `pivot_wider()` function, which also comes from the tidyverse package.

```{r}
# Make sure the tidyverse package is loaded 
data_VO2_wide <- pivot_wider(data = data_VO2_long,
                             names_from = "Time", 
                             values_from = "VO2_max")
head(data_VO2_wide)
```

## Missing Data {-}
Missing data can be handled in many different ways, and selecting the most appropriate method is beyond the scope of this book. Instead, we'll only cover how to find and omit observations with missing data. 

First we'll create a data frame in R that contains `NA` values:

```{r}
data <- data.frame(Col1 = c(NA, 1, 21, 34, NA), 
                   Col2 = c(23, 34, 34, 12, 56),
                   Col3 = c(NA, 2, 12, 43, 12))
data
```

You can use the `is.na()` function to assess if a value is missing in a data frame or not. For example: 

```{r}
is.na(data)
# Missing values in data frame
```

What you get in return is `TRUE` and `FALSE` values. If the value is `TRUE`, that means there is an `NA` value in that location. This might not be particularly helpful, though, if you have a large data frame. Instead, it might be more useful to have the count of `NA` values returned. To do this, we could use the `is.na()` function wrapped inside of a `sum()` function: 

```{r}
sum(is.na(data))
# Sum of missing values in data frame
```

You might want to know how many `NA` values are in a specific column rather than the entire dataset. Selecting just one column from a data frame is a very common task, and one way you can do this is with the dollar sign, `$`:

```{r}
sum(is.na(data$Col1))
# Sum of missing values in Col1 
```

A quick way to get the missing values for all columns of a data frame is with the `colSums()` function:

```{r}
colSums(is.na(data))
```

You can omit all missing values in a data frame with the `na.omit()` function:

```{r}
na.omit(data)
```

The "opposite" of the `na.omit()` function would be finding and listing all the rows that *do* contain missing values, and that's what the code below accomplishes:

```{r}
data[!complete.cases(data), ]
```

A couple things need to be explained first to understand the code above. First, brackets, `[ ]`, are used to subset in R. For example, if you wanted the 5th row and 3rd column returned from a data frame, you could do this by subsetting the data frame like so: `data[5,3]`. The first item within the bracket refers to the row number, and the second number refers to the column number. If you don't put any numbers within the brackets, `data[,]`, then *everything* is selected (the entire data frame). Going back to the example above, you can see that the `data` is being subsetted by the rows. `complete.cases()` is a function similar to `na.omit()`, where the `NA` values are not included. Notice that there is an exclamation point, `!`, in front of the function. This means that instead of returning the cases that are complete, what's being returned is the cases that are *not* complete from the data frame. So, we're subsetting `data` by its rows, where the *incomplete* cases are returned, and all the columns are selected. If that was confusing, that's okay! If there's one thing to take away from this, it's that you can use the brackets for subsetting. 

> Note: You might have noticed that you can select a column in R like this `data$Col3`, and also like this `data[ ,3]`. There are almost always multiple ways to accomplish something in R, and both ways work equally well.

Let's try computing the mean for the first column of the `data` object which contains missing values:

```{r}
mean(data$Col1)
```

As you can see, `NA` is returned. This is because `Col1` contains missing values. Luckily the `mean()` function has an `na.rm` argument, which stands for 'not available, remove'. We can set this argument to `TRUE`, so that missing values will be removed when computing the mean: 

```{r}
mean(data$Col1, na.rm = TRUE)
```

### Mean imputation {-}
Rather than omitting missing values from analysis with the `na.omit()` function, you could perform mean inputation.

```{r}
df <- data.frame(col1 = c(2.5, 4.2, 3.2, NA))
df$col1[is.na(df$col)] <- mean(df$col, na.rm = TRUE)
df
```

In the code above, a data frame is created that has one column, and the fourth value is an `NA`. The `NA` is then replaced with the average of the other three numbers. 

> We are not suggesting that you use mean imputation when your dataset contains missing values, but merely demonstrating how this can be done in R. 

## Recoding Variables {-}
One of the easiest ways to recode/fix factors in R is with the `factor()` function. We'll use the **SubID_Age_Sex.csv** dataset that was used earlier:

```{r}
SubID_Age_Sex <- read.csv("SubID_Age_Sex.csv")
head(SubID_Age_Sex)
```

For the `factor()` function, the first argument, `x`, is the column that should be selected, which in this case is `sex`. Again, we can use the dollar sign `$` to select a specific column from a dataset. The second argument, `levels`, is the levels that should be included, and the third argument, `labels`, is the labels that should be assigned to each level, respectively. 

```{r}
SubID_Age_Sex$sex <- factor(x = SubID_Age_Sex$sex, 
                            levels = c("M","F"), 
                            labels = c(0,1))
head(SubID_Age_Sex)
```

Let's take a look at a recoding example that's slighly more involved. We'll use the **data_VO2.csv** dataset used earlier.

```{r}
data_VO2 <- read.csv("data_VO2_long.csv")
head(data_VO2)
```

And we'll add a column of the subject's sex: 

```{r}
data_VO2$Sex <- rep(c(rep("M", 4), rep("F", 4)), 10)
head(data_VO2)
```

Rather than having VO2max as continuous data, we'll change it to categorical data: `poor`, `fair`, `good`, `excellent`. The catch, however, is that the classification will partially depend on whether the subject was a male or female (it also depends on age, but we'll assume everyone in the study was 30 years old). 

The easiest way to do this is with the `ifelse()` function, and we'll also use several functions from the tidyverse package. First, we need to filter the data so that only males or females are included in the dataset, but not both. We can do this with the `filter()` function from the tidyverse package. We'll also need to create a new column to store the data in, and one easy way to do this is with the `mutate()` function from the tidyverse package.  

```{r}
data_VO2_F <- filter(data_VO2, Sex == "F")

data_VO2_F <- 
mutate(data_VO2, VO2_max_cat = ifelse(VO2_max < 22.8, "Very Poor",
                              ifelse(VO2_max >= 22.8 & VO2_max <= 26.9, "Poor", 
                              ifelse(VO2_max >= 27.0 & VO2_max <= 31.4, "Fair",
                              ifelse(VO2_max >= 31.5 & VO2_max <= 35.6, "Good",
                              ifelse(VO2_max >= 35.7 & VO2_max <= 40.0, "Excellent",
                                     "Superior"))))))
head(data_VO2_F)
```

In the code above, the data was filtered to only include females. Notice that there are two equals signs `==` within the `filter()` function. If you only add one equals sign here you'll get an error, and errors like this can be very difficult to find! After filtering, the data frame was assigned to the object `data_VO2_F`. Next, the `mutate()` function was used to create a new column called `VO2_max_cat` (cat for category). Then, the `ifelse()` function was used to assign a category based on the subject's VO2_max. The first line reads, "If the VO2_max is less than 22.8, assign it the value "Very Poor"." The last category, `Superior`, is assigned if the value does not meet any of the other conditions, which in this case is all values greater than 40.0. 

Now we can repeat the same process for males:

```{r}
data_VO2_M <- filter(data_VO2, Sex == "M") 

data_VO2_M <- 
mutate(data_VO2_M, VO2_max_cat = ifelse(VO2_max < 31.5, "Very Poor",
                             ifelse(VO2_max >= 31.5 & VO2_max <= 35.4, "Poor", 
                             ifelse(VO2_max >= 35.5 & VO2_max <= 40.9, "Fair",
                             ifelse(VO2_max >= 41.0 & VO2_max <= 44.9, "Good",
                             ifelse(VO2_max >= 45.0 & VO2_max <= 49.4, "Excellent",
                                    "Superior"))))))

head(data_VO2_M)
```

If you wanted the data frames to be combined back into one data frame that included both males and females, you could use the `rbind()`  function. 

You might want to recode certain values as `NA` if you know the values are either impossible or very unrealistic. Here's an example were `Age` data is collected, and values less than 0 or greater than 120 are recoded as `NA`. Notice the vertical bar in the code. The vertical bar (also called the pipe) means `or`. So the code below reads, "If `Age` is less than 0 or greater than 120, assign it the value `NA`."  

```{r}
df <- data.frame(Age = c(1546, 43, 23, 56, -64))
df$Age[df$Age < 0 | df$Age > 120] <- NA
df
```

## Changing Variable Types {-}
```{r eval=FALSE}
class()
as.numeric()
as.integer()
as.character()
as.factor()
```

## Rename Columns {-}
```{r eval=FALSE}
colnames()
```

## Change Column Order {-}
```{r eval=FALSE}
# Reorder by column name
data <- data[c("A", "B", "C")]

# Reorder by column index
df[,c(1,2,3,4)]
```

## Data Transformation {-}
```{r eval=FALSE}
select()
filter()
mutate()
arrange()
group_by()
summarise()
```

## Dealing with Dates {-}