# Multiple Regression 

## The `lm()` Function {-}

```{r, eval=FALSE}
lm(formula, data, subset, weights, na.action,
   method = "qr", model = TRUE, x = FALSE, y = FALSE, qr = TRUE,
   singular.ok = TRUE, contrasts = NULL, offset, ...)
```

The `lm()` (linear model) function is used for fitting linear models. This is the function we'll be using for this chapter. Notice that there are many arguments for this function, but the `formula` argument is the only argument that *needs* to be specified. If you'd like to learn more about functions and arguments, please have a look at **Chapter 2** which covers basic programming concepts, including functions and arguments.

# Quick Example {-}
In this example we'll use the `lm()` function for fitting a multiple regression model with two continuous predictors. This will be a quick and simple example that assumes the data is in tidy format and is ready for analysis. If this is not the case for you, you can take a look at **Chapter 4** which covers basic data formatting.

### Model 1 {-} 
$$y_i\ =\ \beta_0\ +\ \beta_1\left(x_i\right)\ +\ \beta_2\left(x_i\right)\ +\ e_i$$

# Comprehensive Example 
```{r eval=FALSE}
# Import
## Set your working directory to the data folder
getwd()
setwd()
list.files()
data <- read.csv()
## You can use the read.csv() arguments to make importing adjustments

## Bind together files if needed 
cbind() # Column bind
rbind() # Row bind

## Long and Wide formatting

# View 
head()
str()
summary()
hist()
plot()
is.na()
mean()
sd()

# Format 

# Model
lm()
```
