```{r include=FALSE}
library(plyr)
library(car)
```

# One-Way ANOVA

> In this chapter we'll use the `lm()` and `ezANOVA()` functions for fitting a one-way ANOVA model with dummy codes and contrast codes, which will take the following form: 

$$y_i=\beta_0+\beta_1\left(X1_L\right)+\beta_2\left(X1_Q\right)+\epsilon_i$$

> The code in this chapter only works if you're following along with the Github folder for this book (which you can download [here](https://github.com/JasonDude16/Statistical-Analysis-Folder)), you've correctly set your working directory to the data folder (which you can learn how to do in [Chapter 4](#CH4)), and run the code in the order it appears in this chapter. 

## Importing {-} 
For this chapter we'll be using the **data_800m.csv** file. This dataset contains information about a subject's VO2max and their 800 meter run time. 

```{r}
data <- read.csv("data_800m.csv")
```

## Viewing {-}
The **data_800m.csv** dataset contains information about a subject's VO2max and their 800 meter run time. Base off of that run time, they were categorized as `faster`, `medium`, or `slower`.

```{r}
head(data)
```

> More examples of viewing data can be found in [Chapter 5](#CH5)

## Formatting {-}
There's no formatting that *needs* to be done for this dataset. By default, R will create dummy codes for categorical variables. Lambda 1 and Lambda 2 will be assigned the codes `0,1,0` and `0,0,1`. However, we could create a set of contrast codes to analyze linear and quadratic effects. We'll recode the `faster`, `medium`, and `slower` categories as `-1`, `0`, `1` and `0`, `1`, `2` for the linear and quadratic codes, respectively. One easy way to do this is with the `mapvalues()` function which comes from the `plyr` package. You'll first need to install the package (`install.packages(plyr)`) and then load the library (`library(plyr)`) before using this function. 

```{r}
data$lin.c <- mapvalues(x = data$groups, 
                        from = c("faster", "medium", "slower"),
                        to = c(1,0,-1))

data$quad.c <- mapvalues(x = data$groups, 
                         from = c("faster", "medium", "slower"),
                         to = c(-1,2,-1))

```

> More examples of formatting data can be found in [Chapter 6](#CH6)

## Modeling {-}

### The `ezANOVA()`Function {-}

```{r, eval=FALSE}
ezANOVA(
    data
    , dv
    , wid
    , within = NULL
    , within_full = NULL
    , within_covariates = NULL
    , between = NULL
    , between_covariates = NULL
    , observed = NULL
    , diff = NULL
    , reverse_diff = FALSE
    , type = 2
    , white.adjust = FALSE
    , detailed = FALSE
    , return_aov = FALSE
)
```

The `ezANOVA()` function is used for ANOVA models. There are many arguments for this function, but not all of them need to be specified. The ones that do need to be specified will depend on the exact ANOVA that you're performing. If you'd like to learn more about functions and arguments, [Chapter 2](#CH2) covers basic programming concepts, including functions and arguments.

### The `lm()` Function {-}

```{r, eval=FALSE}
lm(formula, data, subset, weights, na.action,
   method = "qr", model = TRUE, x = FALSE, y = FALSE, qr = TRUE,
   singular.ok = TRUE, contrasts = NULL, offset, ...)
```

The `lm()` (linear model) function is used for fitting linear models. Notice that there are many arguments for this function, but the `formula` and `data`arguments are the only ones that *need* to be specified. If you'd like to learn more about functions and arguments, [Chapter 2](#CH2) covers basic programming concepts, including functions and arguments.

### Model with Dummy Codes {-}
By default, the `lm()` and `ezANOVA()` functions use the dummy codes `0`, `1`, `0` and `0`, `0`, `1` for three categories. In the formatting section we created contrast codes, but in this example we will not use those contrast codes. We'll let the function create dummy codes for us. 

In this example we'll use the `ezANOVA()` function. We need to specify the `data`, `dv`, `wid`, `between`, and `type` arguments. You can also set `detailed` and `return_aov` to `TRUE` for a more comprehensive output. The `data` argument is set equal to the entire dataset, the `dv` argument, which is short for dependent variable, is set equal to the column of the dependent variable. The `wid` argument is set equal to the column that contains the variable specifying the case/Ss identifier; usually the `subID` column, or equivalent. The `between` argument is set equal to the between-subjects factor(s), which in this case is the `groups` column.

```{r}
ez::ezANOVA(data = data, 
            dv = VO2, 
            wid = subID, 
            between = .(groups),
            type = 3,
            detailed = TRUE, 
            return_aov = TRUE)
```

Notice the formatting in the code above. When using the `between` and `within` arguments, you need to use a period and parentheses, `.()`, when listing the column. You can list additional columns for both of these arguments by separating each column with a comma: `between = .(groups, Column2)`.

### Model with Contrast Codes {-}
Rather than using the default dummy codes, we can use the set of contrast codes that were created in the formatting section above. We'll use the `lm()` function to implement this model. 

```{r}
my_model <- lm(formula = VO2 ~ lin.c + quad.c, data = data)
summary(my_model)
```

```{r}
Anova(my_model, type = "III", singular.ok = T)
```

For comparison purposes, here's what the model printout looks like with the dummy codes using the `lm()` function: 

```{r}
summary(lm(VO2 ~ groups, data = data))
```
