```{r include=FALSE}
library(tidyr)
library(ez)
```

# Mixed-Factorial ANOVA

> In this example we'll use the `ezANOVA()` function for fitting a Mixed-Factorial ANOVA model with one crossed (within-subject) factor, and one nested (between-subjects) factor, which will take the following form: 

$$y_i=\beta_0+\beta_1\left(X1\right)+\beta_2\left(X2\right)+\beta_3\left(X1\cdot X2\right)+U_i+\epsilon_{ij}$$

> The code in this chapter only works if you're following along with the Github folder for this book (which you can download [here](https://github.com/JasonDude16/Statistical-Analysis-Folder)), you've correctly set your working directory to the data folder (which you can learn how to do in [Chapter 4](#CH4)), and run the code in the order it appears in this chapter. 

## Importing {-}
We'll be using the **flexbility_power.csv** file for this example. 
```{r}
data <- read.csv("flexibility_power.csv")
```

## Viewing {-}
In this dataset, male and female subjects performed a countermovement vertical jump under three conditions: `Baseline`, `Post_run`, and `Post_stretch`, where the `Post_run` and `Post_stretch` conditions consisted of running and stretching (lower body) for 10 minutes, respectively, just prior to the vertical jump. Half of the students completed the `Post_run` first and vice versa. The subjects were instructed to try to jump as high as possible under each condition. Vertical jump peak power was then calculated as a function of jump height and body mass.

```{r}
head(data)
```

The data reported under the `Baseline`, `Post_run`, and `Post_stretch` columns is vertical jump peak power, in watts. 

> More examples of viewing data can be found in [Chapter 5](#CH5)

## Formatting {-}
The data should be converted to long format to make it compatible with the `ezANOVA()` function. We can use the `pivot_longer()` function from the `tidyverse` package. Make sure you have that package loaded before using this function.

```{r}
data <- pivot_longer(data = data, 
                     cols = c("Baseline", "Post_run", "Post_stretch"),
                     names_to = "Condition",
                     values_to = "Power_Watts")
```

> More examples of formatting data can be found in [Chapter 6](#CH6)

## Modeling {-}

### The `ezANOVA()`Function {-}

```{r, eval=FALSE}
ezANOVA(
    data
    , dv
    , wid
    , within = NULL
    , within_full = NULL
    , within_covariates = NULL
    , between = NULL
    , between_covariates = NULL
    , observed = NULL
    , diff = NULL
    , reverse_diff = FALSE
    , type = 2
    , white.adjust = FALSE
    , detailed = FALSE
    , return_aov = FALSE
)
```

The `ezANOVA()` function is used for ANOVA models. There are many arguments for this function, but not all of them need to be specified. The ones that do need to be specified will depend on the exact ANOVA that you're performing. If you'd like to learn more about functions and arguments, [Chapter 2](#CH2) covers basic programming concepts, including functions and arguments.

### Model {-}
In this example we need to specify the `data`, `dv`, `wid`, `within` and `between` arguments. The `data` argument is set equal to the entire dataset, the `dv` argument, which is short for dependent variable, is set equal to the column of the dependent variable. The `wid` argument is set equal to the column that contains the variable specifying the case/Ss identifier; usually the `subID` column, or equivalent. The `within` and `between` arguments are set equal to the within-subject and between-subjects factor(s). 

```{r}
ezANOVA(data = data, 
        dv = Power_Watts, 
        wid = ID, 
        within = .(Condition),
        between = .(Sex))
```

Notice the formatting in the code above. When using the `between` and `within` arguments, you need to use a period and parentheses, `.()` when listing the column. You can list additional columns for both of these arguments by separating each column with a comma: `within = .(Condition, Column2)`.