```{r include=FALSE}
library(car)
```

# Analysis of Covariance (ANCOVA)

> In this example we'll use the `aov()` function for fitting one and two sample t-test models with one continuous predictor, which will take the following form: 

$$y_i=\beta_0+\beta_1\left(X1\right)+\beta_2\left(X2\right)+\epsilon_i$$

> The code in this chapter only works if you're following along with the Github folder for this book (which you can download [here](https://github.com/JasonDude16/Statistical-Analysis-Folder)), you've correctly set your working directory to the data folder (which you can learn how to do in [Chapter 4](#CH4), and run the code in the order it appears in this chapter. 

## Importing {-}
```{r}
data <- read.csv("power.csv")
```

## Viewing {-}
In this dataset, males and females performed a maximal effort Wingate Test, which is a 30 second test usually performed on a stationary bike that's used to assess anaerobic leg power. The participant's maximum achieved power, `WG_power_watts`, mass, `mass_kg`, and several other variables were recorded. 

```{r}
head(data)
```

```{r}
ggplot(data, aes(x = mass_kg, y = WG_power_watts, color = sex)) +
    geom_point() + 
    geom_smooth(method = "lm", se = FALSE) + 
    theme_bw() + 
    scale_color_brewer(palette = "Dark2") +
    labs(title = "Wingate Power vs Mass in Males and Females") + 
    xlab("Mass (kg)") +
    ylab("Wingate Power (Watts)")
```

> More examples of viewing data can be found in [Chapter 5](#CH5)

## Modeling {-}

### The `aov()`Function {-}

```{r, eval=FALSE}
aov(formula, data = NULL, projections = FALSE, qr = TRUE,
    contrasts = NULL, ...)
```

The `aov` function has several arguments, but the only ones that need to be specified are the `formula` and `data` arguments. 

### Model {-}
When using the `aov()` function, the `formula` argument is set equal to the dependent variable, followed by a tilde, `~`, and then the independent variable(s). The `data` argument is set equal to the object that contains the dataset, which in this example is the object called `data`.

```{r}
my_model <- aov(formula = WG_power_watts ~ sex + mass_kg, data = data)
```

You can use the `summary.lm()` function to see a summary of the model. 

```{r}
summary.lm(my_model)
```